---
title: "Music Hit Prediction - CMSC 320 Final Project"
author: "Jacob Luterman, Joshua Pooranmal, Shiv Patel"
date: "05/18/2020"
output: html_document
---

Music has the power to culturally, morally, and emotionally influence our society. As time progresses, and popular culture inevitably shifts, diffrent genres of music begin to surface and become popular. Whether it's 1940s swing, 1970s rock-and-roll, or 1990s hip-hop, society's ear for music is constantly changing. In order for record labels (both old and new) to maintain relevance and continue to be proftable, it is important to have some way of keeping up with contemporary society's musical preferences. Playing it by ear is seems like a good idea, but this introduces a lot of subjectivity, which isn't reliable. Tracking user feedback is also a good idea if you're looking to gauge how well a song is currently doing, but the point is to stay ahead of the game, not alongside it. 

So how does one work to predict whether or not a song will be a hit? Is this even possible? 

The following project code makes use of data collected from Spotify that represents all songs from the 1960s up until the 2010s. For each song entity, we have access to the values of several compositional factors (tempo, energy, loudness, etc.), its artist(s), and whether or not the song is considered a "hit" by Spotify. The goal of this project is to collect and evaluate trends in the makeup of songs as time progresses in order to train a machine learning model that can predict if a song will be a hit based solely on its compositional makeup. In order to evaluate if our models are accurate, we will train them using all songs from 1960s-2000s and identify how accurately it can predict whether or not a 2010s song is a hit.

## Data curation and cleaning

The first step of the process is to curate our data. We will combine all data from the 1960s to the 2010s into a single data frame. We will also append a new column called 'decade' to identify which decade a song belongs to for later use. 

```{r results='hide', message=FALSE, warning=FALSE}
library(rvest)
library(tidyverse)
library(lubridate)
library(tidymodels)
library(caret) # Confusion matrix, algorithm training
library(mice) # Basic data preprocessing - eg. Removing null values(na.omit)etc
library(ggplot2) # All plots
library(ggcorrplot) # All correlation plot
library(dplyr) # For data manipulation .. eg. selecting numeric variables)
library(openxlsx) # read excel file
library(knitr)
options(stringsAsFactors = FALSE)
```


```{r Importing Data}
# Datasets retrieved from https://www.kaggle.com/theoverman/the-spotify-hit-predictor-dataset
songs_2010 <- "dataset-of-10s.csv"
songs_2000 <- "dataset-of-00s.csv"
songs_1990 <- "dataset-of-90s.csv"
songs_1980 <- "dataset-of-80s.csv"
songs_1970 <- "dataset-of-70s.csv"
songs_1960 <- "dataset-of-60s.csv"

song_files <- c(songs_2010,songs_2000,songs_1990,songs_1980,songs_1970,songs_1960)
song_df <- NULL
for (song_file in song_files) {
  tmp_df <- read.csv(song_file)
  decade <- str_extract(song_file,"\\d{2}")
  tmp_df$decade <- decade
  song_df <- bind_rows(song_df,tmp_df)
}
song_df$decade <- ordered(song_df$decade, levels = c("60","70","80","90","00","10"))

as_tibble(song_df)
```

Once our data frame is generated, we must get rid of data that is not important for the purpose of this project. We don't need the resource url of a song so we can remove the uri column. Although a track's key does speak to its composition, most music listeners aren't aware of the key they are listening to, so we will omit this. We also decided to get rid of valence because we felt like measuring "positiveness" of a song introduces some subjectivity and we weren't interested in the potential overhead of including this attribute in our models. Below is the song dataframe that will be used for the majority of this project:
```{r  Tabling DataFrame}
song_df <- song_df %>% 
  select(-c(uri,key,valence))

as_tibble(song_df)
```

The "target" attribute identifies whether or not a song is a hit by assigning hits with the attribute walue '1' and non-hits with '0'. Here is a data frame of all songs Spotify considers a hit:
```{r Only Hits Table}
hits_only <- filter(song_df,target == 1) 

as_tibble(hits_only)
```

## Exploratory Data Analysis

Now that we have our working dataframe, we can try and identify changes in specific compositional aspects of hits and non-hits across decades. 

```{r Graphing Tempo vs. Decade}
plot_avg_by_decade <- function(df,col){
  
plotting <- df %>% 
  group_by(decade,target) %>% 
  summarise(avg=mean({{col}})) %>% 
  mutate(target=as.factor(target))

  ggplot(plotting) + 
    geom_bar(stat="identity", aes(x=decade,y = avg,fill=target), position = "dodge") +
    labs(title = "Tempo Changes per Decade Between Hits(1) and Non-Hits(0)", x = "Decade", y = "Average Tempo") +
    theme(plot.title = element_text(hjust = 0.5)) 
}

song_df %>% 
  plot_avg_by_decade(tempo)
```

Tempo is the overall estimated tempo of a track in beats per minute (BPM). In musical terminology, tempo is the speed or pace of a given piece and derives directly from the average beat duration. The graph above conveys that there is not much change in average tempo as the decades change. Furthermore, the tempo between hits(blue) and non-hits(red) aren't too different as well for each decade. This indicates that listeners' taste for tempo seldom changes. Is this because humans are inherently drawn to listen to songs within a given tempo range?

```{r Graphing Energy vs. Decade}
plot_avg_by_decade <- function(df,col){
  
plotting <- df %>% 
  group_by(decade,target) %>% 
  summarise(avg=mean({{col}})) %>% 
  mutate(target=as.factor(target))

  ggplot(plotting) + 
    geom_bar(stat="identity", aes(x=decade,y = avg,fill=target), position = "dodge") +
    labs(title = "Energy Changes per Decade Between Hits(1) and Non-Hits(0)", x = "Decade", y = "Average Energy") +
    theme(plot.title = element_text(hjust = 0.5)) 
}

song_df %>% 
  plot_avg_by_decade(energy)
```

Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy. Average energy seems to have a more obvious trend then that of energy. Within each decade, songs that were hits had higher average energy ratings. Average energy also seems to grow for both hits and non-hits as we move from decade to decade. There is a slight drop from the 2000s to 2010s, but nothing that would overwhelmingly indicate a future trend towards lower energy songs. 

```{r Graphing Acousticness vs. Decade}
plot_avg_by_decade <- function(df,col) {
  
  plotting <- df %>% 
    group_by(decade,target) %>% 
    summarise(avg=mean({{col}})) %>% 
    mutate(target=as.factor(target))
  
  ggplot(plotting) + 
    geom_bar(stat="identity", aes(x=decade,y = avg,fill=target), position = "dodge") +
    labs(title = "Acousticness Changes per Decade Between Hits(1) and Non-Hits(0)", x = "Decade", y = "Average Acousticness") +
    theme(plot.title = element_text(hjust = 0.5)) 
}

song_df %>% 
  plot_avg_by_decade(acousticness)
```

Acoustiness is a confidence measure from 0.0 to 1.0 of whether the track is acoustic. 1.0 represents high confidence the track is acoustic. Within each decade, hits tended to have lower acoustic ratings than non-hits. This trend persists as time progresses as average acousticness for both hits and non hits sees a drop in each decade. It seem like listeners' are not too interested in songs with high acoustiness ratings. 

```{r Liveliness vs. Decade}
plot_avg_by_decade <- function(df,col){
  
  plotting <- df %>% 
    group_by(decade,target) %>% 
    summarise(avg=mean({{col}})) %>% 
    mutate(target=as.factor(target))
  
  ggplot(plotting) + 
    geom_bar(stat="identity", aes(x=decade,y = avg,fill=target), position="dodge") +
    labs(title = "Liveness Changes per Decade Between Hits(1) and Non-Hits(0)", x = "Decade", y = "Average Liveliness") +
    theme(plot.title = element_text(hjust = 0.5)) 
}

song_df %>% 
  plot_avg_by_decade(liveness)
```

Liveness Detects the presence of an audience in the recording. Higher liveness values represent an increased probability that the track was performed live. A value above 0.8 provides strong likelihood that the track is live. Liveness originally seems to indicate that there isn't much change, but if you look closely, hits subtly decrease in liveness as time progresses. Contemporary listeners may be more interested in studio-recorded music than they are in live music.  

```{r Durations vs. Decade}
plot_avg_by_decade <- function(df,col){
  
  plotting <- df %>% 
    group_by(decade,target) %>% 
    summarise(avg=mean({{col}})) %>% 
    mutate(target=as.factor(target))
  
  ggplot(plotting) + 
    geom_bar(stat="identity", aes(x=decade,y = avg,fill=target), position ="dodge") +
    labs(title = "Duration(ms) Changes per Decade Between Hits(1) and Non-Hits(0)", x = "Decade", y = "Average Duration(ms)") +
    theme(plot.title = element_text(hjust = 0.5)) 
}

song_df %>% 
  plot_avg_by_decade(duration_ms)
```

Duration is measured in milliseconds. Listeners seemed to prefer longer songs going into the 1990s, but subsequently began to grow a renewed liking to more concise pieces.

```{r Time Signature vs. Decade}
plot_avg_by_decade <- function(df,col){
  
  plotting <- df %>% 
    group_by(decade,target) %>% 
    summarise(avg=mean({{col}})) %>% 
    mutate(target=as.factor(target))
  
  ggplot(plotting) + 
    geom_bar(stat= "identity",aes(x=decade,y = avg,fill=target),position = "dodge") +
    labs(title = "Time Signature Changes per Decade Between Hits(1) and Non-Hits(0)", x = "Decade", y = "Average Time Signature") +
    theme(plot.title = element_text(hjust = 0.5)) 
}

song_df %>% 
  plot_avg_by_decade(time_signature)
```

Time signature is an estimated overall time signature of a track. The time signature (meter) is a notational convention to specify how many beats are in each bar (or measure). Time signatures seem to stay fairly consistent between both hits and non-hits as time progresses.

```{r Chorus vs. Decade}
plot_avg_by_decade <- function(df,col){
  
  plotting <- df %>% 
    group_by(decade,target) %>% 
    summarise(avg=mean({{col}})) %>% 
    mutate(target=as.factor(target))
  
  ggplot(plotting) + 
    geom_bar(stat="identity", aes(x=decade,y = avg,fill=target), position="dodge") +
    labs(title = "Chorus Hit Changes per Decade Between Hits(1) and Non-Hits(0)", x = "Decade", y = "Average Chorus Hit") +
    theme(plot.title = element_text(hjust = 0.5)) 
}

song_df %>%
  plot_avg_by_decade(chorus_hit)
```

Chorus hit is the artist's best estimate of when the chorus would start for the track. Its the timestamp of the start of the third section of the track (in milliseconds). There is a general trend for all songs: all choruses for both hits and non-hits seems to hit for the first time around the 40 second mark of a song. It's quite intriguing that no decade has worked to break this trend. 

```{r Sections vs. Decade}
plot_avg_by_decade <- function(df,col){
  
  plotting <- df %>% 
    group_by(decade,target) %>% 
    summarise(avg=mean({{col}})) %>% 
    mutate(target=as.factor(target))
  
  ggplot(plotting) + 
    geom_bar(stat="identity", aes(x=decade,y = avg,fill=target), position = "dodge") +
    labs(title = "Sections Changes per Decade Between Hits(1) and Non-Hits(0)", x = "Decade", y = "Average Sections") +
    theme(plot.title = element_text(hjust = 0.5)) 
}

song_df %>% 
  plot_avg_by_decade(sections)
```

Sections refers to the number of sections the particular track has. Similar to duration statistics, it seems like listeners yearned for more sections in their music up until the 2000s, when hits started to become relatively more concise. 

```{r Loudness vs. Decade}
plot_avg_by_decade <- function(df,col){
  
  plotting <- df %>% 
    group_by(decade,target) %>% 
    summarise(avg=mean({{col}})) %>% 
    mutate(target=as.factor(target))
  
  ggplot(plotting) + 
    geom_bar(stat="identity", aes(x=decade,y = avg,fill=target), position="dodge") +
    labs(title = "Loudness Changes per Decade Between Hits(1) and Non-Hits(0)", x = "Decade", y = "Average Loudness") +
    theme(plot.title = element_text(hjust = 0.5)) 
}

song_df %>% 
  plot_avg_by_decade(loudness)
```

The overall loudness of a track in decibels (dB). Loudness values are averaged across the entire track and are useful for comparing relative loudness of tracks. Loudness is the quality of a sound that is the primary psychological correlate of physical strength (amplitude). Values typical range between -60 and 0 db. As time has progessed, it seems like music as a whole (both hits and non-hits) tend to become more loud. This may be due to the use of more technology when producing songs. Electronic devices can emulate sounds that are loud better than concrete instruments themselves. An example would be the 808 drum in hip-hop music.

```{r Speechiness vs. Decade}
plot_avg_by_decade <- function(df,col){
  
  plotting <- df %>% 
    group_by(decade,target) %>% 
    summarise(avg=mean({{col}})) %>% 
    mutate(target=as.factor(target))
  
  ggplot(plotting) + 
    geom_bar(stat= "identity",aes(x=decade,y = avg,fill=target),position = "dodge") +
    labs(title = "Speechiness Changes per Decade Between Hits(1) and Non-Hits(0)", x = "Decade", y = "Average Spechiness") +
    theme(plot.title = element_text(hjust = 0.5)) 
}

song_df %>% 
  plot_avg_by_decade(speechiness)
```

Speechiness detects the presence of spoken words in a track. The more exclusively speech-like the recording (e.g. talk show, audio book, poetry), the closer to 1.0 the attribute value. From the 60s to the 80s, rock-and-roll was very popular, which is why hits tended to have less spechiness. With the rise of modern pop and hip-hip, hits are now far more speechy. There is little evidence to assert this this upward trend in speechiness will soon come to an end. 

```{r Instrumentalness vs. Decade}
plot_avg_by_decade <- function(df,col){
  
  plotting <- df %>% 
    group_by(decade,target) %>% 
    summarise(avg=mean({{col}})) %>% 
    mutate(target=as.factor(target))
  
  ggplot(plotting) + 
    geom_bar(stat="identity", aes(x=decade,y = avg,fill=target), position="dodge") +
    labs(title = "Instrumentalness Changes per Decade Between Hits(1) and Non-Hits(0)", x = "Decade", y = "Average Instrumentalness") +
    theme(plot.title = element_text(hjust = 0.5)) 
}

song_df %>% 
  plot_avg_by_decade(instrumentalness)
```

Instrumentalness predicts whether a track contains no vocals. “Ooh” and “aah” sounds are treated as instrumental in this context. Rap or spoken word tracks are clearly “vocal”. The closer the instrumentalness value is to 1.0, the greater likelihood the track contains no vocal content. There is an abvious trend that listeners lean towards songs that are more vocal than those that are instrumental. This preference has gotten even more obvious as time has progressed, with hits for the 2010s not even surpassing an average instrumentalness rating of 0.05.

## Machine Learning

With our newfound info about the compositional makeup of hits and non-hits between decades, we can now try and build a machine learning model that can predict wether or not a song is a hit. As mentioned previously, we will train our model using all songs up to and including the 2000s. We will then use our model to predict whether or not a 2010s song will be a hit. We will compare our model's result (hit or non-hit) with the actual target value from Spotify to measure the model's accuracy.

### Random Forest

The first model we will use will be a random forest. The code below filters specific entities and using tidymodel, we build our testing and training data. We also preporcoess our data to ensure that it can be processed correctly before building our model.

Training data: all songs from 1960s to 2000s, inclusive
Test data: all songs from 2010s

```{r}
song_df %>% 
  select(-c(artist,track,mode,sections,time_signature)) %>% 
  mutate(target = as.factor(target))-> update_song_df
  training_data <- update_song_df %>% 
    filter(decade!="10")
  testing_data <- update_song_df %>% 
     filter(decade=="10")
data_recipe <- training_data %>% recipe(target~.) %>% 
  step_corr(all_numeric()) %>% 
  step_center(all_numeric(),-all_outcomes()) %>% 
  step_scale(all_numeric(),-all_outcomes()) %>% 
  prep()

data_testing <- data_recipe %>% 
  bake(testing_data)
data_training <- juice(data_recipe)

```


Now that we have curated our training and testing data, we can build a random forest model. We will use 100 trees for this model. Once done, we will use this model to predict whether or not a songs from the 2010s will be a hit.

```{r}
r_forest <- rand_forest(trees = 100, mode = "classification") %>% 
  set_engine("randomForest") %>% 
  fit(target~., data = data_training)

r_forest %>%
  predict(data_testing) %>% 
  bind_cols(data_testing) -> predictions_r_forest
  predictions_r_forest %>% 
    metrics(truth = target,estimate = .pred_class)
```
Our machine learning model was able to predict hits at roughly a 79% success rate. Using this model, record labels, as long as they have the correct compositional metrics for a song, can predict if it will be a hit with roughly 80% confidence. Not bad at all.   

### Multiple Logistic Regression

```{r}
mus = read.csv("song_df.csv") #read song dataset to variable mus

musiclog = mus[,-c(1,2,3,4)] #removing the  X ,track,artist,uri as these cannot be modelled

summary(musiclog) #summary of dataset
```

Here we just tidy our dataframe to include only attributes that we can train our model on. We remove attributes that cannot be modelled on.


#### Correlation Analysis

```{r Correlation Analysis}

musiclognum = musiclog %>% 
  select_if(is.numeric) # selecting only the numeric variables from student for correlation

#Correlation plot from numeric variables
ggcorrplot(cor(musiclognum), 
           hc.order = TRUE, 
           outline.col = "white",
           lab = TRUE, 
           title = "Correlation Plot", 
           insig = c("pch", "blank"), 
           pch = 4, 
           pch.col = "black",
           lab_size = 2,
           tl.cex =8)
```

The correlation plot helps showcase the relationship between numeric variables. instrumentalness variable has the highest correlation (.41) with the Target variable, followed by danceability.

#### Logistic Regression Model

```{r Logistic Regression Model}
musiclog$decade = as.factor(musiclog$decade)#converting decade variable to factor
set.seed(123)#used to randomize the records in the dataset

train_set <- musiclog %>% filter(decade != "0" & decade != "10" )#training set obtained by filtering on data below 2000
train_set <-train_set[,-c(17)]

test_set <- filter(musiclog, decade == "0" | decade == "10")#testing set obtained by filtering on data from 2000 till now
test_set <-test_set[,-c(17)]

logitreg <- glm(target~.,data = train_set, family = "binomial")#Creating a logistic regression model based on all the independent variables
  
#Model summary
summary(logitreg)

predicttrain <- predict(logitreg,test_set,type = "response")#generating predicted output as a probability


Predicted_Result <- as.factor(ifelse(predicttrain > .6,1,0))#checking if predicted output is greater than 60% threshold and then assigning the target class(1/0)

#confusion matrix b/w predicted output and test data output
confusionMatrix(table(Predicted = Predicted_Result,Actual = test_set$target),positive = "1")
```


A logistic regression model was created based on all the independant variables and the dependant variable Target(1/0).  
From the model summary it is clear that at 95% confidence interval all the independendant variables except - valence,time_signature and duration_ms were statistically significant in determining the Target(1/0). This was confirmed as the variables had p-value less than 0.05 thus rejecting the null hypothesis that coefficient = 0.

By keeping the cutoff for the Target variable to be considered as hit to .6 , a confusion matrix was created. The confusion matrix shows the model has an accuracy of 74.56% which is greater than the No - information rate. The confusion matrix also shows the sensitivity(82.46%) and specificity(66.6%) for the model.


#### Variable Importance

```{r Variable Importance}

#Creating a data frame showing variable importance in decreasing order of importnace based on T-statistic
imp <- as.data.frame(varImp(logitreg))
imp <- data.frame(Variable_names   = rownames(imp),overall_Importance_Tstat = imp$Overall)
imp[order(imp$overall,decreasing = T),]

```

Our model identifies instrumentalness, danceability, and speechiness as the most important variables when determining a hit. This makes sense. Many songs nowadays are low in instrumentalness and high in speechiness. Furthermore, (pre-pandemic) contemporary society is big on clubs and social gatherings, so danceability is very important. 

## Future Implications and Stretch Goals

One overarching realization that we considered but did not put into our model was the idea that once an artist has a hit, they are more likely to get another. Our model predicts hits, but not "breakout" hits, or those that came from artist who never previously had a hit. 

```{r Graphing Breakout Hits}
song_df %>% 
  group_by(decade,artist) %>%
  summarise(has_hit=max(target))%>%
  mutate(has_hit=as.factor(has_hit)) %>% 
  group_by(decade,has_hit)%>% 
  count -> hit_amount 

ggplot(hit_amount) + 
  geom_bar(stat="identity", aes(x=decade,y=n,fill=has_hit), position="dodge") +
  labs(title = "Hits per Decade Between Breakout Artists(0) and Artists with Existing Hits(1)", x = "Decade", y = "Number of Hits") +
  theme(plot.title = element_text(hjust = 0.5)) 
```

The graph above shows that it used to be far more difficult to get a "breakout" hit. The growth of technology has given aspiring artists a plotform where they can broadcast and promote their own music and brad, which may be the reason why the numbers begin to even out as time progresses. Our model doesn't consider the technological aspect of society. Songs can go viral at any time. Take, for instance, Old Town Road by Lil Nas X. The song blew up on social media before it became the longest reigning #1 hit on Billboard top 100. Maybe instances like this are impossible to predict, but still something worth noting nonetheless. 

## Reference Material
Project data source:
https://www.kaggle.com/theoverman/the-spotify-hit-predictor-dataset

Another music hit predictor:
https://towardsdatascience.com/song-popularity-predictor-1ef69735e380

Request more song data from Spotify:
https://developer.spotify.com/documentation/web-api/reference/tracks/get-audio-features/